---
title: "ariadne: integrate prior knowledge in your bioinformatic analysis"
author:
  - name: Giulio Benedetti
    affiliation: University of Turku
    email: giulio.benedetti@utu.fi
output:
    BiocStyle::html_document:
        fig_height: 7
        fig_width: 10
        toc: true
        toc_float: true
        toc_depth: 2
        number_sections: true
        self_contained: true
        code_folding: show
# bibliography: ../inst/REFERENCES.bib
date: "`r doc_date()`"
package: "`r pkg_ver('ariadne')`"
vignette: >
  %\VignetteIndexEntry{ariadne: integrate prior knowledge into your bioinformatic analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

## Motivation

All domains of biology involve relational information, such as links between
genes, organisms, and their interactions. For example, organisms form symbiotic
or antagonistic relationships, consume specific foods, or avoid others, creating
a web of connections that define biological systems. When this relational
information is systematically organised and made accessible through databases
such as UniProt (proteins), Rhea (reactions), and KEGG (metabolic pathways), it
empowers scientists to build upon decades of research, effectively standing "on
the shoulders of giants." However, leveraging these resources in routine
bioinformatic analyses is often hindered by challenges such as inaccessible data
repositories, incompatible formats, and the sheer volume of data that exceeds
local storage capacities. Without a structured approach, navigating the
labyrinth of biological knowledge can feel overwhelming, time-consuming, and
discouraging.

ariadne is designed to simplify this process by "Assembling Relational
Information Across the Database Network," offering a harmonised graph of
interlinked biological resources and a robust toolkit to explore it. This system
is build upon a reproducible pipeline, implemented in the companion package
ariadneDB, where the harmonised graph is constructed and maintained with its
release history available in Zenodo.

ariadne can be used for knowledge-driven annotation, feature stratification,
and goal-oriented selection. For instance, it enables functional annotation of
microbial features using databases like ChocoPhlAn or WoL, and stratifying them
by Gut Modules or Bug Signatures. It can also extend to other biological data
types, such as metabolites, proteins, or genes, making it a flexible tool for
integrating relational knowledge into routine analyses.

## Implementation

| **Term** | **Definition** | **Examples** |
|----------|----------------|--------------|
| **Feature type** | A distinct type of biological entity with unique elements labeled by IDs. Written in lowercase for consistency. | `genes` (KEGG genes ID), `ko` (KEGG Orthologues), `ec` (Enzyme Commission number), `uniref90` (UniRef sequence clusters) |
| **Level** | A unique element belonging to a feature type. Often referred to as a *feature*. | 563 (NCBI ID for *E. coli*), C00246 (KEGG compound ID for butyrate), P0DTC2 (UniProtKB ID for SARS-CoV-2 spike protein) |
| **Linkmap** | A 2-column data frame with feature types as columns and their links in the rows. Named as `x2y`, where x and y are feature types. | `tax2cpd` (linkmap between taxon names and compounds) |
| **Resource** | A database from which linkmaps between feature types can be retrieved. | KEGG, Rhea, UniProt |
| **Weave** | The operation of sequentially linking feature types to create a new linkmap between two previously unlinked feature types. Uses the syntax `x ~ y`, where x is the origin and y is the target feature type. | `ko ~ ec`, `disease ~ drug`, `genes ~ metacyc` |

Main strengths of this approach:

- **Graph-based resource representation** using an igraph object where edges
  are resources and nodes are feature types, enabling graph theory for path
  selection and easy expansion with custom resources.
- **Fast queries**: with httr2 for API requests and SPARQL for even faster
  linkmap retrieval when possible.
- **Efficient caching and storage**: to automatically cache linkmaps with
  BiocFileCache and store them in the efficient Apache Parquet format via arrow.
- **Optimised linkmap generation** through the MultiFactor class, which
  streamlines factor operations to create linkmaps by sequentially combining
  them along the path from origin to target feature type.

# Tutorial

## Installation

Release:

```{r install-release, eval = FALSE}
if( !requireNamespace("BiocManager", quietly = TRUE) ){
    install.packages("BiocManager")
}

BiocManager::install("ariadne")
```

Beta:

```{r install-beta, eval = FALSE}
remotes::install_github("Minotau-R/ariadne")
```

## Import and explore the resource graph

ariadne combines different resources into a single graph highlighting the links
between biological entities. The resource graph can be imported as an igraph
object with the function `ariadne`.

The way that ariadne retrieves and combines relational data is guided by its
companion data package [ariadneDB](https://github.com/Minotau-R/ariadneDB),
which contains information about the feature types and mappings available from
a handful of popular databases.

The relational information of each resource is described by a graph, where nodes
and edges represent feature types and mappings between them, respectively. In
ariadneDB, these graphs are stored in the Graph Modelling Language (GML), which
is a standard format for relational data.

```{r lib, message = FALSE, warning = FALSE}
library(ariadne)
library(igraph)
library(dplyr)
library(ggplot2)
```

When `ariadne` is executed, the graph for each resource is retrieved from the
database release on [Zenodo](https://zenodo.org/records/18788725) and combined
into a single igraph object.

```{r fetch}
# Import resource graph
graph <- ariadne()
```

The graph can also be visualised with `plotPath`.

```{r plot}
# Plot resource graph
plotPath(graph)
```

Let's say we need to find all links between gene functions and enzyme activity
represented by KEGG Orthologues (KO) and Enzyme Commission (EC) ids,
respectively. The function `searchPath` can be used to find the `k` shortest
paths in the graph.

```{r}
# Search for paths from ko to ec
searchPath(graph, ko ~ ec, k = 3)
```

```{r}
plotPath(graph, ko ~ ec, include = "uniref90", prune = TRUE, focus = TRUE)
```

The graph can be explored using all native functions from the igraph package.
For example, we can sort its nodes by degree, that is, the number of adjacent
edges:

```{r}
node_degrees <- graph |>
    degree(mode = "all") |>
    sort(decreasing = TRUE)

node_degrees <- node_degrees[node_degrees >= 10]

knitr::kable(node_degrees, col.names = "degree")
```

In addition, we can list all the nodes in the neighbourhood of a certain node:

```{r}
neighbors(graph, "ko", mode = "all")
```

## ariadneDB and resource versioning

Some resources have multiple versions, which can be selected through the
argument `versions`. This defaults to the latest version found in the database.

```{r res_versions}
# List available resource versions
versions <- listResourceVersions(default = TRUE)

# View versions
knitr::kable(versions)
```

The igraph object is made of two components: edges data and nodes (or vertices)
data, which are stored in two different data tables.

```{r graph2df}
# Convert igraph to data.frame pair
graph_df <- igraph::as_data_frame(graph, what = "both")

# Extract edges data
edge_df <- graph_df$edges
# Extract nodes data
node_df <- graph_df$vertices
```

As mentioned above, the resource graph is composed of feature types (nodes) and
mappings (edges) from several resources. Here are the summary statistics for
each resource integrated within ariadne:

```{r stats}
# Get stats for each resource
df <- edge_df |>
    group_by(source) |>
    summarise(edges = n(), nodes = n_distinct(c(from, to)))

# View stats
knitr::kable(df)
```

The graph for each resource can be visualized with its own graph.

```{r plot_graphs}
# Create resource graphs
plots <- lapply(
    df$source,
    function(x) plotPath(graph, res.name = x, layout = "stress", focus = TRUE) +
        ggtitle(x)
)

# View resource graphs
plots
```

## Weaving with weavePath vs weaveComplex

We decide to go through the simplest path, which corresponds to a `k` of 1. The
links between KO and EC are found by the function `weavePath`, which fetches
the necessary resources and weaves a path from KO to EC.

```{r}
# Weave path from ko to ec
ko2ec <- weavePath(graph, ko ~ ec, k = 1)

# View some links
head(ko2ec)
```

The links are provided as a **linkmap**, a 2-column table where each row is a
pair of ids (one KO and one EC). When available, the feature names are returned
as a third optional column.

## Exploring results

```{r}
# x2y
```

```{r}
# plotModules
```

## Bioconductor SummarizedExperiment integration

```{r}
# tse <- addModules(tse, tax2path)
```

```{r}
# agglomerateByModule
```

```{r}
# applyByModule
```

```{r}
# viz functional alpha/beta
```

# Resources

## Citation

We hope that ariadne will be useful for your research. Please use the following
information to cite the package and the overall approach. Thank you!

```{r citation}
## Citation info
citation("ariadne")
```

## Acknowledgements

ariadne originates from the joint effort of the R/Bioconductor community. It is
mainly based on the following software:

- [R](https://www.r-project.org/), statistical programming language [@core2025r]
- [MultiFactor](https://github.com/thomazbastiaanssen/anansi/),
  Annotation-based Analysis of Specific Interactions [@bastiaanssen2023anansi]

## Help

You can reach us by one of the communication channels listed
[here](https://microbiome.github.io/OMA/docs/devel/pages/support.html). We are
happy to receive questions, suggestions as well as contributions. For the last
point, check the
[contributor guidelines](https://microbiome.github.io/mia/CONTRIBUTING.html).

# Reproducibility

R session information:

```{r reproduce, echo=FALSE}
# Session info
options(width = 120)
sessionInfo()
```
